---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2024"
subtitle: "Assignment 2 - Due date 02/25/24"
author: "Zhenghao Lin"
output: pdf_document
geometry: margin=2.54cm
---

## Submission Instructions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github.   

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp24.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## R packages

R packages needed for this assignment:"forecast","tseries", and "dplyr". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here
library("forecast")
library("tseries")
library("dplyr")
library("ggplot2")
```


## Data set information  

Consider the data provided in the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx" on our **Data** folder. The data comes from the US Energy Information and Administration and corresponds to the December 2023 Monthly Energy Review. The spreadsheet is ready to be used. You will also find a $.csv$ version of the data "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source-Edit.csv". You may use the function $read.table()$ to import the $.csv$ data in R. Or refer to the file "M2_ImportingData_CSV_XLSX.Rmd" in our Lessons folder for functions that are better suited for importing the $.xlsx$.

```{r}
#Importing data set
getwd()
energy_data <- read.table(file="./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.csv", header=TRUE,dec = ".",sep=",",stringsAsFactors = TRUE) 
```


## Question 1
You will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. Use the command head() to verify your data.

```{r}
#New data frame
energy_data <- energy_data %>% select("Total.Biomass.Energy.Production", "Total.Renewable.Energy.Consumption", "Hydroelectric.Power.Consumption")

#Verification
head(energy_data)
```


## Question 2
Transform your data frame in a time series object and specify the starting point and frequency of the time series using the function ts().

```{r}
ts_energy <- ts(energy_data[,1:3], start = c(1973, 1), frequency = 12)
ts_energy
```


## Question 3
Compute mean and standard deviation for these three series.

```{r}
#Mean and standard deviation for total biomass energy production
biomass_total_mean <- mean(ts_energy[,1])
biomass_total_mean
biomass_total_stdv <- sd(ts_energy[,1])
biomass_total_stdv
#Mean and standard deviation for total renewable energy production
renewable_total_mean <- mean(ts_energy[,2])
renewable_total_mean 
renewable_total_stdv <- sd(ts_energy[,2])
renewable_total_stdv
#Mean and standard deviation for hydroelectric power consumption
hydro_mean <- mean(ts_energy[,3])
hydro_mean
hydro_stdv <- sd(ts_energy[,3])
hydro_stdv
```


## Question 4
Display and interpret the time series plot for each of these variables. Try to make your plot as informative as possible by writing titles, labels, etc. For each plot add a horizontal line at the mean of each series in a different color.

```{r}
autoplot(ts_energy[,1]) + 
  geom_hline(yintercept = biomass_total_mean, color = "blue") +
  xlab("Time") +
  ylab("Total Biomass Energy Production") +
  labs(color="Total.Biomass.Energy.Production") +
   ggtitle("Time Series of Total Biomass Energy Production")

autoplot(ts_energy[,2]) + 
  geom_hline(yintercept = renewable_total_mean, color = "blue") +
  xlab("Time") +
  ylab("Total Renewable Energy Production") +
  labs(color="Total.Renewable.Energy.Production") +
   ggtitle("Time Series of Total Renewable Energy Production")

autoplot(ts_energy[,3]) + 
  geom_hline(yintercept = hydro_mean, color = "blue") +
  xlab("Time") +
  ylab("Hydroelectric Power Consumption") +
  labs(color="Hydroelectric.Power.Consumption") +
   ggtitle("Time Series of Hydroelectric Power Consumption")
```


## Question 5
Compute the correlation between these three series. Are they significantly correlated? Explain your answer.

```{r}
cor(ts_energy[,1:3])
```
The total biomass energy production is highly postive-correlated to total renewable energy consumption. Hydroelectric power consumption, however, is slightly negatively correlated to total biomass energy production and barely has any correlation with total renewable energy consumption.

## Question 6
Compute the autocorrelation function from lag 1 up to lag 40 for these three variables. What can you say about these plots? Do the three of them have the same behavior?

```{r}
Acf(ts_energy[,1], lag.max = 40, main = "ACF for Total Biomass Energy Production")
Acf(ts_energy[,2], lag.max = 40, main = "ACF for Total Renewable Energy Production")
Acf(ts_energy[,3], lag.max = 40, main = "ACF for Hydroelectric Power Consumption")
```
For total biomass energy production and total renewable energy consumption, within 40 lags, the graph displays a slowly-decaying trend, indicating that there exists a trend in its time series data.
For hydroelectric power consumption, within 40 lags, not only can we observe a slowly-decaying trend from the graph, but we can also discover an obvious seasonality from the graph of this time series, indicating that such time series data has both trend and seasonality.

## Question 7
Compute the partial autocorrelation function from lag 1 to lag 40 for these three variables. How these plots differ from the ones in Q6?

```{r}
Pacf(ts_energy[,1], lag.max = 40, main = "PACF for Total Biomass Energy Production")
Pacf(ts_energy[,2], lag.max = 40, main = "PACF for Total Renewable Energy Production")
Pacf(ts_energy[,3], lag.max = 40, main = "PACF for Hydroelectric Power Consumption")
```
Unlike the ACFs, in PACFs of the three time series data, all three graphs seem to cut off at certain lag after several significant lags.

